---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "",
  message = FALSE,
  warning = FALSE,
  out.width = "100%", 
  dpi = 300
)


# Packages
library(tidyverse)
library(targets)
```

# NERACOOS_CPR_DATA

Integrating Continuous Plankton Recorder data into NERACOOS & ERDDAP

This repository contains the necessary code and documentation to support hosting the [Gulf of Maine Continuous Plankton Recorder Survey Data](https://www.fisheries.noaa.gov/feature-story/long-running-plankton-survey-resume-gulf-maine) on [ERDDAP](http://ismn.erddap.neracoos.org/erddap/info/index.html?page=1&itemsPerPage=1000).

## Organization:

This repository documents the data provenance for continuous plankton recorder data obtained from a number of scientific research agencies ([NOAA](https://www.fisheries.noaa.gov/about/northeast-fisheries-science-center) and [MAB](https://www.mba.ac.uk/)), and covering different sampling transects (The Gulf of Maine & The Mid-Atlantic Bight Transects).

Raw data from all sources is contained in the `data_raw/` directory. Code that prepares the raw data for ERDDAP and any necessary documentation is specific to the source that the data was received from. This information can be found in the following sub-folders:

| Sub-Folder             | Description                                    |
| ---------------------- | ---------------------------------------------- |
| GulfOfMaine_NOAA       | Gulf of Maine CPR Data obtained from NOAA      |
| GulfOfMaine_MBA        | Gulf of Maine CPR Data Obtained from MBA       |
| MidAtlantic_NOAA       | Mid-Atlantic Bight CPR Data Obtained from NOAA |
| MidAtlantic_MBA        | Mid-Atlantic Bight CPR Data Obtained from MBA  |


These resources have been processed independently due to differences in measurement units and organization structures. Documentation on how each dataset was received and treated prior to uploading into ERDDAP is documented within each of the corresponding sub-folders.



### Reproducing the Data Transformations


The full processing pipeline from the raw data to their final ERDDAP formats has been implemented using the [{targets}](https://docs.ropensci.org/targets/) R-package, and can be recreated in full by running the following code in an active R session. (Assuming all R-packages are installed). 

```
library(targets)
tar_make()
```

This will recreate the processing steps outlined in `_targets.R` that transform the raw files into the format uploaded onto ERDDAP:

```{r gulf of maine pipeline}
# #tar_visnetwork()
# withr::with_dir(rprojroot::find_root('_targets.R'), 
#                 tar_glimpse(allow =  c("GOM_NOAA_RAW", "zooplankton_raw", "zoo_abundances", "zoo_key", "noaa_zooplankton_erddap", "save_noaa_zooplankton")))


withr::with_dir(rprojroot::find_root('_targets.R'), 
                tar_visnetwork(allow =  c("GOM_NOAA_RAW", "zooplankton_raw", "zoo_abundances", "zoo_key", "noaa_zooplankton_erddap", "pivot_zooplankton", "pull_zoo_pieces")))
```

The DAG above shows a simplified representation of the steps for the NOAA ZPR Zooplankton data, where the taxonomic information found in the header is separated from the abundance information and later joined back after it has been reshaped. Similar cleanup paths exist for the data obtained from NOAA as well as the data obtained from the MBA. 

### Unit and Taxa Classification Differences

Due to how the CPR data is stored and maintained within these two institutions, conversions to a standard unit of measurement is necessary when working with CPR jointly from both sources. In addition to unit conversions, there are taxonomic and development stages that are recorded inconsistently across them, which require the use of a key for transitioning to coarser scale groupings.


Information on resolving the differences between these two data resources can be found in the following sub folder: `working_across_sources/`, with examples of code working from ERDDAP as a starting point.




# Project Funding:

Funding for making these resources available was provided through grant awards from the [National Science Foundation](https://www.nsf.gov/) and from the [Lenfest Ocean Program](https://www.lenfestocean.org/en). With communication and support from the Northeast Fisheries Science Center and the Marine Biological Association.


---

## Bonus Resources:



