---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "",
  message = FALSE,
  warning = FALSE,
  out.width = "100%", 
  dpi = 300
)


# Packages
library(tidyverse)
library(targets)
```


# Gulf of Maine CPR Transect Data (NOAA Sourced)


## About:

This folder contains documentation on the Gulf of Maine CPR data obtained directly from NOAA, and adjustments made prior to its integration into ERDDAP.


## NOAA Gulf of Maine CPR Data

Data in its original form can be found in the [data_raw/](www.github.com/gulfofmaine/continuous_plankton_recorder/data_raw) sub-directory. Data in its original form was obtained from the NOAA Northeast Fisheries Science Center through communication with Chris Melrose. 

#### Data Acquisition:

Data obtained from NOAA/NEFSC spans the period of: 1961-2013, and was delivered in the following excel file:

> **data_raw/NOAA_1961-2013/Gulf of Maine CPR (Feb 14, 2014 update).xlsx**

#### Data Organization:

That excel file is divided into 2 sheets. One contains data on phytoplankton and the other contains records on zooplankton. Each excel sheet has two or more additional rows in the header explaining the identification and development stage of certain taxa. 

**NOTE: CPR data from NOAA is reported in units that do not follow CPR survey conventions. (ref: Jossi et al. 2003)**


| Sheet Number  | Description                                    | Units               |
| ------------- | ---------------------------------------------- | ------------------- |
| Sheet 1       | Phytoplankton taxa densities by silk transect  |  Abundance / 1m3    |
| Sheet 2       | Zooplankton taxa densities by silk transect    |  Abundance / 100m3  |



#### Data Processing Code

To make the CPR data more "ERDDAP-friendly" minor changes were applied to reshape the structure of the original dataset, resulting in a longer & [tidy-er dataset](https://r4ds.had.co.nz/tidy-data.html). Any modifications to the original files have been documented in the R scripts held in the [R/](www.github.com/gulfofmaine/R) sub-directory and more explicitly detailed in the steps of `_targets.R`. 

Taxonomic encodings from the original excel file headers are also checked against MARMAP codes to ensure accuracy https://www.nefsc.noaa.gov/nefsc/Narragansett/taxcodesA.html


## ERDDAP Metadata


To make the datasets available on ERDDAP, appropriate metadata had to be detailed as a .xml file: `erddap_xml/dataset.xml`. This file populates the metadata information for the two datasets [zooplankton dataset_id](http://ismn.erddap.neracoos.org/erddap/tabledap/noaa_gom_traverse.html) & [phytoplankton dataset_id](http://ismn.erddap.neracoos.org/erddap/tabledap/noaa_gom_phyto.html). The information in the `datasets.xml` file is displayed online via the erddap interface, and is what allows ERDDAP to process data filtering services when accessing/downloading the data. 

An additional pdf with history on the survey and key references is also available in `refs/gom_cpr_metadata.pdf`. This folder also contains additional references on the history of the survey.


---

## Working with CPR Data

Data can be accessed directly from [ERDDAP](http://ismn.erddap.neracoos.org/erddap/index.html), or through software packages like R's {[rerddap](https://github.com/ropensci/rerddap/tree/0723cafce46cf31dd46efed71d76706788c08889)}.

```{r, eval = FALSE, echo = TRUE}
# Accessing the Gulf of Maine CPR Zooplankton Data
library(rerddap)

# Get the tabledap information from the server link and dataset_id
cpr_info <- info(url = "http://ismn.erddap.neracoos.org/erddap", 
                 datasetid = "noaa_gom_cpr_zooplankton")

# Use the tabledap function to import all the data (optionally add filters)
gom_zooplankton <- tabledap(cpr_info)

```


Or python's {[erddapy](https://ioos.github.io/erddapy/)}.

```{python, eval = FALSE, echo = TRUE}
# Accessing the Gulf of Maine CPR Zooplankton Data
from erddapy import ERDDAP
e = ERDDAP(
  server = "http://ismn.erddap.neracoos.org/erddap",
  protocol = "tabledap",
  response = "csv"
)

# Add Dataset ID
e.dataset_id = "noaa_gom_cpr_phytoplankton"

# Coerce to pandas df
gom_phytoplankton = e.to_pandas()

```



